---
title: "ESM 244 Lab Week 5"
author: "Casey O'Hara"
output: html_document
---

### Attach required packages:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)


# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(palmerpenguins)
library(patchwork)

# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs

```

# Part 1: Principal components analysis (PCA)

Principal components analysis is an ordination method allowing us to glean as much about our multivariate data as possible in a simplified number of dimensions.

Here, we'll use the `penguins` data within the `palmerpenguins` R package to explore variable relationships and clustering by species in a PCA biplot. For this example with PCA, we will only use the structure size measurements (bill length and depth, mass, and flipper length).

```{r}
penguin_clean <- penguins %>%
  drop_na()

penguin_pca <- penguin_clean %>%
  dplyr::select(body_mass_g, ends_with("_mm")) %>%
  scale() %>%
  prcomp()


autoplot(penguin_pca,
     	data = penguin_clean,
     	loadings = TRUE,
     	colour = 'species',
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5
     	) +
  scale_color_manual(values = c("blue","purple","orange")) +
  scale_fill_manual(values = c("blue","purple","orange")) +
  theme_minimal()

# It's not perfect, but it's enough for now...
```

``` {r}
# Variance explained by each PC
screeplot(penguin_pca, type = "lines")
screeplot(penguin_pca, type = "barplot")

# See the loadings (weighting for each principal component)
penguin_pca$rotation

```

## Biplot loadings by hand


```{r}
loadings_df <- data.frame(penguin_pca$rotation) %>%
  mutate(axis = row.names(.))

new_pts_df <- data.frame(penguin_pca$x)

ggplot() +
  geom_point(data = new_pts_df, aes(x = PC1, y = PC2), color = 'blue') +
  geom_segment(data = loadings_df,
               x = 0, y = 0, aes(xend = PC1, yend = PC2, group = axis),
               arrow = arrow(length = unit(.25, 'cm'))) +
  geom_text(data = loadings_df,
            aes(x = PC1, y = PC2, label = axis), vjust = 0, nudge_y = .01) +
  theme_minimal()
```
## Screeplot by hand

```{r}
sd_vec <- penguin_pca$sdev
var_vec <- sd_vec^2 ### standard deviation is sqrt of variance!
pc_names <- colnames(penguin_pca$rotation)

pct_expl_df <- data.frame(v = var_vec / sum(var_vec),
                          pc = fct_inorder(pc_names)) %>%
  mutate(pct_lbl = paste0(round(v*100, 1), '%'))

ggplot(pct_expl_df, aes(x = pc, y = v)) +
  geom_col() +
  geom_text(aes(label = pct_lbl), vjust = 0, nudge_y = .002) +
  labs(x = 'Principal component', y = 'Variance explained')
```



# Part 2: `ggplot` customization & reading in different file types

We spent some time in ESM 206 customizing our data visualizations. Let's add some more tools, including:
- Highlight spaghetti plots with `gghighlight`
- An interactive graph with `plotly`

Here, we'll also read in stored .txt and .xlsx, and files from a URL to build our toolkit for how to read in data.

**Data:** NOAA [Foreign Fisheries Trade Data](https://www.fisheries.noaa.gov/national/sustainable-fisheries/foreign-fishery-trade-data)

#### Read in a .xlsx file, & do some wrangling

```{r}
fish_noaa <- read_excel(here("data","foss_landings.xlsx")) %>%
  clean_names() %>%
  mutate(across(where(is.character), tolower)) %>% # convert all characters to lowercase
  mutate(nmfs_name = str_sub(nmfs_name, end = -4)) %>%  # remove last 4 characters
  filter(confidentiality == "public")
```

Now, let's make and customize a graph:

```{r}
fish_plot <- ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  theme_minimal()

# Make it interactive:
ggplotly(fish_plot)

# Highlight series based on condition(s):
ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line() +
  gghighlight(nmfs_name == "tunas") + # Highlight just tunas
  theme_minimal()

ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  gghighlight(max(pounds) > 1e8) + # Highlight just tunas
  theme_minimal()
```

#### Read in data from a URL, `lubridate()` refresher, `mutate()` as a workhorse to transform variables

- See paletteer color palettes:
- Discrete with `View(palettes_d_names)`
- Continuous with `View(palettes_c_names)`

**Data:** Monroe Water Treatment Plant Daily Electricity Use

Accessed from [data.gov](https://catalog.data.gov/dataset/monroe-water-treatment-plant-energy-and-water-consumption/resource/5afc8aa7-b485-4173-bcc6-c56270efedb8)

Google: data.gov Monroe Water Treatment Plant Daily Electricity Use (probably the first result)

Summary: "Daily energy use (kWh), demand (kW), and volume water treated (million gallons). 2010 through current. A second electric meter and account were added at the plant in March 2013. The usage and demand data from this meter are labeled as "Energy Use 2" and "Peak 2."

The URL to the CSV file is provided at the website above (or copy from below):

```{r}
monroe_wt <- read_csv("https://data.bloomington.in.gov/dataset/2c81cfe3-62c2-46ed-8fcf-83c1880301d1/resource/13c8f7aa-af51-4008-80a9-56415c7c931e/download/mwtpdailyelectricitybclear.csv") %>%
  clean_names()

monroe_ts <- monroe_wt %>%
  mutate(date = mdy(date)) %>% # Convert date to a stored date class
  mutate(record_month = month(date)) %>% # Add column w/ month number
  mutate(month_name = month.abb[record_month]) %>% # Add column w/ month abbreviation
  mutate(month_name = fct_reorder(month_name, record_month)) # Make month name a factor & reorder based on values in record_month column

ggplot(data = monroe_ts, aes(x = month_name, y = total_k_wh)) +
  geom_jitter(aes(color = month_name),
          	show.legend = FALSE,
          	alpha = 0.5,
          	size = 0.3,
          	width = 0.2)
```

PART 1: INTERACTIVE RMARKDOWN
---
title: 'Part 1: Interactive RMarkdown'
author: "Allison Horst"
output: html_notebook
runtime: shiny
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```
 
To create this document, make sure to choose the "Shiny document" option when you create the file! That will add the `runtime: shiny` line to your YAML, and give you some nice examples you can check out.
 
Once you create the new RMarkdown Shiny document, click 'Run Document' up top to see what appears. An interactive document where users can change what they see!
 
How cool is this for your boss, client, etc? Some of the benefits of an interactive app, without the bulk -- this is a nice option for simple & sleek documents where you want the user to have some flexibility about variables, etc. that appear in a graph or table.
 
Let's make something new.
 
### Attach packages
 
In the setup chunk, attach:
 
- `tidyverse`
- `palmerpenguins`
 
### Add some interactivity!
 
```{r flippers, echo=FALSE}
inputPanel(
  selectInput(inputId = "pick_spp", label = "Pick penguin species:",
          	choices = c("Adélie" = "Adelie", "Chinstrap", "Gentoo"), selected = "Adelie"),
 
  sliderInput("pt_size", label = "Set point size:",
          	min = 1, max = 10, value = 5, step = 0.5)
)
 
penguin_set <- reactive({
  penguins %>%
  filter(species == input$pick_spp)
})
 
renderPlot({
  ggplot(data = penguin_set(), aes(x = flipper_length_mm, y = body_mass_g)) +
	geom_point(size = input$pt_size)
})
```
 
...which is pretty cool.
 
You can publish to a free [shinyapps.io](https://www.shinyapps.io/) account!

 

 
PART 3: SHINY THEMING WITH shinythemes & bslib
 
#
# This is a Shiny web application. You can run the application by clicking
# the 'Run App' button above.
#
# Find out more about building applications with Shiny here:
#
#    http://shiny.rstudio.com/
#
 
library(shiny)
library(bslib)
 
# In console, run bs_theme_preview() to play around with different things!
 
# See ?bs_theme() for more options & information.
 
my_theme <- bs_theme(
  bg = "pink",
  fg = "purple",
  primary = "yellow",
  base_font = font_google("Delius")
)
 
# Define UI for application that draws a histogram
ui <- fluidPage(theme = my_theme,
 
	# Application title
    titlePanel("Old Faithful Geyser Data"),
 
	# Sidebar with a slider input for number of bins
	sidebarLayout(
    	sidebarPanel(
            sliderInput("bins",
                        "Number of bins:",
                  	  min = 1,
                        max = 50,
                        value = 30)
    	),
 
    	# Show a plot of the generated distribution
    	mainPanel(
           plotOutput("distPlot")
    	)
	)
)
 
# Define server logic required to draw a histogram
server <- function(input, output) {
 
	output$distPlot <- renderPlot({
    	# generate bins based on input$bins from ui.R
    	x	<- faithful[, 2]
    	bins <- seq(min(x), max(x), length.out = input$bins + 1)
 
    	# draw the histogram with the specified number of bins
    	hist(x, breaks = bins, col = 'darkgray', border = 'white')
	})
}
 
# Run the application
shinyApp(ui = ui, server = server)
