---
title: "ESM 244 Lab Week 5"
author: "Casey O'Hara"
output: html_document
---

### Attach required packages:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)


# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(patchwork)

# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs
```

# Part 1: Principal components analysis (PCA)

Principal components analysis is an ordination method allowing us to glean as much about our multivariate data as possible in a simplified number of dimensions.

Here, we'll use [data from the Human Development Index](https://hdr.undp.org/data-center) (raw data and metadata are saved in the `data` folder in case you'd like to explore in more detail).  This is a selected set of variables:

* Life expectancy, inequality in life expectancy
* Gross per capita national income, inequality in income
* Expected years of schooling, mean years of school
* CO2 production per capita, material footprint (tonnes per capita)
* Latitude

PCA requires continuous numeric data with no NAs.  So we must drop categorical and character data, and exclude any rows with NAs.  We should also rescale so all numeric variables have a mean 0 and sd 1.

```{r}
hdi_data <- read_csv(here('data/hdi_clean.csv')) %>%
  drop_na() %>%
  mutate(hdicode = factor(hdicode, levels = c('Low', 'Medium', 'High', 'Very High')))

# glimpse(hdi_data)

hdi_pca <- hdi_data %>%
  dplyr::select(where(is.numeric)) %>%
  # dplyr::select(-iso3, -country, -hdicode)) %>%
  # dplyr::select(ends_with('2021'), lat) %>%
  scale() %>%
  prcomp()
```


``` {r}
# See the loadings (weighting for each principal component)
hdi_pca$rotation
```

## Biplot loadings by hand


```{r}
loadings_df <- data.frame(hdi_pca$rotation * 8) %>%
  mutate(axis = row.names(.))

new_pts_df <- data.frame(hdi_pca$x)

ggplot() +
  geom_point(data = new_pts_df, aes(x = PC1, y = PC2, color = hdicode), 
             color = 'blue') +
  geom_segment(data = loadings_df,
               x = 0, y = 0, aes(xend = PC1, yend = PC2, group = axis),
               arrow = arrow(length = unit(.25, 'cm'))) +
  geom_text(data = loadings_df,
            aes(x = PC1, y = PC2, label = axis), vjust = 0, nudge_y = .01) +
  theme_minimal()
```

and by autoplot

``` {r}
autoplot(hdi_pca,
     	data = hdi_data,
     	loadings = TRUE,
     	colour = 'hdicode',
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5
     	) +
  scale_color_manual(values = c('red', 'orange', 'yellowgreen', 'darkgreen')) +
  scale_fill_manual(values = c('red', 'orange', 'yellowgreen', 'darkgreen')) +
  theme_minimal()

# It's not perfect, but it's enough for now...
```

## Screeplot by hand

first by hand, then by screeplot
```{r}
sd_vec <- hdi_pca$sdev
var_vec <- sd_vec^2 ### standard deviation is sqrt of variance!
pc_names <- colnames(hdi_pca$rotation)

pct_expl_df <- data.frame(v = var_vec,
                          pct_v = var_vec / sum(var_vec),
                          pc = fct_inorder(pc_names)) %>%
  mutate(pct_lbl = paste0(round(pct_v*100, 1), '%'))

ggplot(pct_expl_df, aes(x = pc, y = v)) +
  geom_col() +
  geom_text(aes(label = pct_lbl), vjust = 0, nudge_y = .002) +
  labs(x = 'Principal component', y = 'Variance explained')
```

```{r}
# Variance explained by each PC
screeplot(hdi_pca, type = "lines")
screeplot(hdi_pca, type = "barplot")
```

-----


# Part 2: `ggplot` customization & reading in different file types

We spent some time in ESM 206 customizing our data visualizations. Let's add some more tools, including:
- Highlight spaghetti plots with `gghighlight`
- An interactive graph with `plotly`

Here, we'll also read in stored .txt and .xlsx, and files from a URL to build our toolkit for how to read in data.

**Data:** NOAA [Foreign Fisheries Trade Data](https://www.fisheries.noaa.gov/national/sustainable-fisheries/foreign-fishery-trade-data)

## Read in a .xlsx file, & do some wrangling

```{r}
fish_noaa <- read_excel(here("data", "foss_landings.xlsx")) %>%
  janitor::clean_names() %>%
  mutate(across(where(is.character), tolower)) %>% # convert all characters to lowercase
  mutate(nmfs_name = str_sub(nmfs_name, end = -4)) %>%  # remove last 4 characters
  filter(confidentiality == "public")
```

Now, let's make and customize a graph:

```{r}
fish_plot <- ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  theme_minimal()

# Make it interactive:
ggplotly(fish_plot)

# Highlight series based on condition(s):
ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line() +
  gghighlight(nmfs_name == "rockfishes") + # Highlight just rockfishes
  theme_minimal()

ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  gghighlight(max(pounds) > 1e8) + # Highlight high-catch
  theme_minimal()
```

## Read in data from a URL, `lubridate()` refresher, `mutate()` as a workhorse to transform variables

- See paletteer color palettes:
- Discrete with `View(palettes_d_names)`
- Continuous with `View(palettes_c_names)`

**Data:** Monroe Water Treatment Plant Daily Electricity Use

Accessed from [data.gov](https://catalog.data.gov/dataset/monroe-water-treatment-plant-energy-and-water-consumption/resource/5afc8aa7-b485-4173-bcc6-c56270efedb8)

Archived here: 

* https://raw.githubusercontent.com/oharac/esm244_w22_lab5/master/data/MWTP_Elec_Daily.csv
* NOTE: can navigate to https://github.com/oharac/esm244_w22_lab5/blob/master/data/MWTP_Elec_Daily.csv then click the "view raw" button to get a downloadable .csv

Summary: "Daily energy use (kWh), demand (kW), and volume water treated (million gallons). 2010 through current. A second electric meter and account were added at the plant in March 2013. The usage and demand data from this meter are labeled as "Energy Use 2" and "Peak 2."

The URL to the CSV file is provided at the website above (or copy from below):

```{r}
monroe_wt <- read_csv("https://raw.githubusercontent.com/oharac/esm244_w22_lab5/master/data/MWTP_Elec_Daily.csv") %>%
  janitor::clean_names()

monroe_ts <- monroe_wt %>%
  mutate(date = lubridate::mdy(date)) %>% # Convert date to a stored date class
  mutate(record_month = lubridate::month(date)) %>% # Add column w/ month number
  mutate(month_name = month.abb[record_month]) %>% # Add column w/ month abbreviation
  mutate(month_name = fct_reorder(month_name, record_month)) # Make month name a factor & reorder based on values in record_month column

ggplot(data = monroe_ts, aes(x = month_name, y = total_kwh)) +
  geom_jitter(aes(color = month_name),
          	show.legend = FALSE,
          	alpha = 0.5,
          	size = 0.3,
          	width = 0.2) +
  theme_classic()
```


